# Getting Started {#intro}



## Using R as a calculator

* You can use R as a basic **calculator**. 

* For example, if we just type in `42 + 17` into the R console, it should print out the sum:

```{r, echo=TRUE}
42 + 17 
```



* We can compute the square root of 243, $1.56^{124}$, and $7.21 \times 8^{4}$, just by typing these
expressions into the **R** console
```{r, echo=TRUE}
sqrt(243)

1.56*124

7.21*8^4
```

## Variables in R

* When working with more complicated mathematical operations in **R**, it is often useful to
store intermediate values in **named variables**.

* For example, 

```{r, echo=TRUE}
x <- (42 + 17)*sqrt(43)
y <- 7.21*8^4 + log(2.34)
z <- x/y
z  ## print out the value of z
```

* Here, `x`, `y`, and `z` are examples of **variables**.

* The pair of characters `<-` used together is known as the **assignment operator** in **R**.
`x <- 2` assigns the value `2` to the variable `x`.

* A **variable** is the **named storage** of a value (or an object) in memory.

* Why do we need variables?

  + To **reuse** the same value later on.
  + To **generalize** an expression to use in many cases.

* How to use variables in **R**?
  
  + To set the value of a variable, use **assignment** operator `<-`
  
  + To use the value, simply use the **variable name** as if it were its stored value.
  
* For example, ...


## Rules for choosing **variable names** in R

* Variables can be named however you want as long as you follow the 
several variable-naming rules that R has. 

* In **R** variable names can include the following:
  + **letters**: A-Z a-z
  + **digits**: 0-9
  + **underscore and period**: _ .

* Additional rules:
  + Variable names **must start with letters or a period (not underscore or digits)**
  + If a variable name starts with a **period**, it **cannot** be followed by a **number**.
  + Variable names are **case sensitive**.

## Examples of **valid** and **invalid** variable names in R
  
```{r, echo=FALSE}
add.varlid <- c("i", "my_variable", "answer42", ".name")
add.invalid <- c("2things", "location@", "_user.name",".3rd")
AddOperators <- data.frame(Valid=add.varlid, Invalid=add.invalid)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 2), padding=2)
```

* While you are free to choose variable names however you like as long as you follow the variable-naming rules of **R**, making variable names **descriptive** is highly **recommended**.

* **Descriptive** variable names make it easier to read code. This is very helpful if:
    + You are sharing your code or
    + Looking back at code you wrote many weeks/months ago

* Using a **consistent** convention for naming variables is recommended, too:

https://r4ds.had.co.nz/workflow-basics.html


## Variable **Assignment**

* Variables can be assigned using either `<-` or `=`

```{r, echo=TRUE}
x = 123    # Use = to assign a variable
y <- 123   # Or use <- to assign a variable
```

```{r, echo=TRUE}
x   # Retrieve the value of x
y   # Retrieve the value of y
```

* The pair of characters `<-` is the classic symbol used for variable assignment in **R**.

* The use of `<-` instead of `=` is often recommended in **R** style guides:
    + http://adv-r.had.co.nz/Style.html


### **<-** vs. **=**

* `<-` and `=` will work the same if they are both
used in the "usual way" (when assigning variables within 
or outside of a function).

* One exception, is when used inside a **function call**. For example, if we use `=` in the function `sd(x)`:
```{r, echo=FALSE}
rm(x)
```

```{r, echo=TRUE}
sd(x = c(1,2,3,4,5)) # only sets the argument x in sd(x) to (1,2,3,4,5)
#x      ## will return an error if we try to print x
```

```{r, echo=TRUE}
sd(x <- c(1,2,3,4,5)) # This actually assigns the vector (1,2,3,4,5) to x
x
```

* However, using something like `sd(x <- c(1,2,3,4,5))` where we assign variables in a function call is not really done that often. 

* It is **not common** to assign variables in a function call (I never do it).

* Whenever, using a function `f` with a keyword such as `x`, you will generally
want to call that function using `f(x = ...)`

* So, in my opinion, there is not really a strong reason to prefer using `<-` over `=` for assignment. 

* There are other justifications for using `<-` such as the ability 
to do assignment from the left by using the reverse symbol `->`
```{r}
c(1, 2, 3, 4) -> a # Using c(1,2,3,4) = a will not work!  
a
```


## **Types** of variables

* Variables can be used to store **different types** of values.

* Common types include numeric, text, and logical values.
```{r, echo=TRUE}
x <- 3.2
x
```

* Here, `x` is actually a **vector** (basically a collection of elements storing the same type of data).

* It is a vector of **length one** (i.e., it only has one element). 

* This is the reason why you see `[1]` printed out next to the number 3.2. 
    + This means that the **first element** of the vector `x` is $3.2$. 

* **R** treats every variable as some type of collection (e.g., vectors, matrices, lists, etc.).
    + There are no separate data types in **R** for individual numbers.


* The elements in a vector can have different **types** (or modes).

* You can find the types of the elements in a vector by using the
function **typeof**
```{r, echo=TRUE}
y <- sqrt(1743)
typeof(y)  # double and integer are the two numeric types 
```

```{r, echo=TRUE}
z <- 3 # R automatically treats every number as double
z
typeof(z)
```

* The other common types for the elements in a vector include
    + **logical** (TRUE or FALSE) values
    + **character** basically text, e.g., "hello", "car", ...
```{r, echo=TRUE}
y <- TRUE
typeof(y)   
```

```{r, echo=TRUE}
z <- "dog" # to define a character variable, place it inside quotes
typeof(z)
```

* We will discuss these types in more detail later on when
we discuss **vectors**, **matrices**, and **lists**.


## R Operations with **numbers**
  
* As we mentioned before, ...  

```{r, echo=FALSE}
add.operator <- c("+", "-", "*", "/", "%%", "^", "**")
add.meaning <- c("addition", "subtraction", "multiplication","division","remainder","exponent","exponent")
add.example <- c("5 + 8", "90 - 10", "4 * 7", "7 / 2","7 %% 2","3 ^ 4","3 ** 4")
add.result <- c("13", "80", "28","3.5","1","81","81")
AddOperators <- data.frame(Operator=add.operator, Meaning=add.meaning, Example =add.example, Result = add.result)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 4), padding=4)
```

* R operations with numbers have similar **precedence rules** to arithmetic operations

```{r, echo=FALSE}
add.operator <- c("+, -", "*, /, %%", "**, ^", "(expressions...)")
add.description <- c("addition and subtraction", "multiplication, division, remainder","exponentiation","Parenthesis")
add.precedence <- c("low", "...", "...","high")
AddOperators <- data.frame(Operator=add.operator, Description=add.description, Precedence = add.precedence)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 3), padding=3)
```

* Examples of operation **precedence** can be seen when typing the following
expressions into the **R** console:

```{r, echo=TRUE}
1 + 2 *3 ^ 4 # power > mult/div > add/sub
(1 + 2 ) *3 ^ 4 # parenthesis > power
```

## Exercises

* Compute the number 
```{=tex}
\begin{equation}
\frac{\sqrt{1.43 + 5^{1.2}}}{3} 
\end{equation}
```
directly in the **R** console.

* Write an **R script** that assigns the value ...

```{=tex}
\begin{equation}
\ln\Big( 1 + \exp(-2^{1.4}) \Big) + \ln\Big(1 + 2\exp(3^{1.7}) \Big)
\end{equation}
```

* ... to a variable named `x` and prints the result in the Console when you run
the script.

* Which of the following is **NOT** a valid variable name in **R**?
    + .independent_variable3
    + _independent_variable3
    + independent_variable3
    + independent.variable3




