# (PART) R Programming Fundamentals {-}

# Logical Expressions and If-Else Statements in R {#ifelse}



## Logical Expression in R

* A **Logical** expression is an expression that evaluates to either **TRUE** or **FALSE**.

* The following are examples of logical expressions in **R**:
    + `4 > 2`
    + `3 <= 5`
    + `15.0 + 1.3*1.3 > 17.0`
    + ` "cat" == "dog" `
    
* Each of the above expressions will evaluate to either
**TRUE** or **FALSE** if you run them in R.

```{r, echo=TRUE}
4 > 2
```

```{r, echo=TRUE}
3 <= 5
```

```{r, echo=TRUE}
15.0 + 1.3*1.3 > 17.0
```

```{r, echo=TRUE}
x <- "cat" == "dog" # assign to the variable x the value 
                    # returned by this logical expression
x
```

---

* Most logical expressions are constructed by using some **combination** of:
    + **Comparison** operators (<, <=, ==, !=)
    + **Logical** operators (**and**, **or**, **not**) (in **R**: &&, ||, !)

### Comparison Operators

```{r, echo=FALSE}
add.operator <- c("<", ">", "<=", ">=", "==", "!=")
add.meaning <- c("Less than", "Greater than", "Less than or equal to","Greater than or equal to","Equal to","Not equal to")
add.example <- c("5 < 3", "5 > 3", "3 <= 6", "4 >= 3","2 == 2","'str' != 'stR'")
add.result <- c("FALSE", "TRUE", "TRUE","TRUE","TRUE","TRUE")
AddOperators <- data.frame(Operator=add.operator, Meaning=add.meaning, Example =add.example, Result = add.result)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 4), padding=4)
```

### Logical Operators

* The first main logical operator we will discuss is the **logical AND**

* In **R**, the logical operator `&` is used to represent the logical **AND**

* The **logical AND** is used to test whether or not two statements are **both** true.

* For two logical expressions **A** and **B**, the logical expression **A & B** is true
only if both **A** and **B** evaluate to true.

```{r, echo=TRUE}
4 > 2 & 5/2 == 1   ## only the first statement is TRUE
4 > 2 & "car" == "truck"   ## only the first statement is TRUE
4 > 2 & 3 < 5   ## both statements are TRUE
```

---


* The logical operator `|` is used in **R** to represent the logical **OR**.

* For two Boolean expressions **A** and **B**, the Boolean expression **A | B** is true if at least one of **A** and **B** evaluates to true.

* Note that if **A** and **B** are both true, **A | B** will be true; **or** does not mean only one of
**A** and **B** is true.

```{r, echo=TRUE}
4 > 2 | 5/2 == 1  ## only the first statement is TRUE
4 > 2 | "car" == "truck"  ## only the first statement is TRUE
4 > 2 | 3 < 5  ## both statements are TRUE
```


---

* The logical operator **!** is used to represent the logical **NOT**.

* If the logical expression **A** is true, then **! A** is false.

```{r, echo=TRUE}
!4 > 2  
!4 > 2 & 3 > 1  ## !4 > 2 is FALSE
!(!4 > 2 & !3 > 1) ## expression in parentheses is evaluated first
```

---

* Note that we can apply logical operations to the keywords `TRUE` and `FALSE` themselves:
```{r, echo = TRUE}

TRUE & FALSE ## logical AND

TRUE | FALSE ## logical OR

!TRUE  ## logical NOT

```

* The below table summarizes the logical operations discussed.
  
```{r, echo=FALSE}
add.operator <- c("!","", "&&", "", "", "", "||", "", "", "")
add.meaning <- c("Logical NOT", "", "Logical AND", "", "", "", "Logical OR", "", "", "")
add.example <- c("!TRUE", "!FALSE", "FALSE & FALSE", "TRUE & FALSE", "FALSE & TRUE", "TRUE & TRUE",
                 "FALSE | FALSE", "TRUE | FALSE", "FALSE | TRUE", "TRUE | TRUE")
add.result <- c("FALSE", "TRUE", "FALSE", "FALSE", "FALSE", "TRUE", "FALSE", "TRUE", "TRUE", "TRUE")
AddOperators <- data.frame(Operator=add.operator, Meaning=add.meaning, Example=add.example, Result = add.result)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 4), padding=4)
```

### Precedence with logical operations

```{r, echo=FALSE}
add.operators <- c("&, |, !", "+, -", "*, /, %%", "**, ^", "(expressions ...)")
add.description <- c("Boolean operators", "Addition and subtraction", "Multiplication, division, remainder", "Exponentiation", "Parenthesis")
add.precedence <- c("Low", "", "","", "High")
AddOperators <- data.frame(Operators=add.operators, Meaning=add.description, Precedence =add.precedence)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 3), padding=3)
```

* Mathematical operations are generally performed before logical operations.
```{r, echo=TRUE}
4 + 2 > 5
4 + 2 == 6
```



### Abbreviating TRUE and FALSE with T and F

* You can use `T` and `F` in place of `TRUE` and `FALSE` 
    + I usually **do not** use `T` and `F`, but you will often see `T` and `F` used.
```{r, echo = TRUE}
T     ## T is shorthand for TRUE

F     ## F is shorthand for FALSE

T && F

T || F
```


* While you can use `T` and `F` in place of `TRUE` and `FALSE`, it is good practice to be **careful** when using these logical **abbreviations**.

```{r, echo = TRUE}

T <- 2   ## T is defined as a variable

          ## Now T represents a vector with 
T         ## a number value, not TRUE

F        ## F still represents FALSE

## FALSE <- 3   # this will result in an error

## TRUE and FALSE cannot be redefined, thus safer to use
```


### Examples of logical operations in R

```{r, echo = TRUE}

TRUE | FALSE  ## boolean OR

!TRUE          ## NOT operator

!TRUE | TRUE    ## Which one will get evaluated first?

! (TRUE | TRUE) ## How about this time?
```

### logical operators and vectors

* You can also apply the logical operators `&`, `|`, `!` to two vectors.

* As an example, let us first define two logical vectors `x` and `y` of length 4
```{r, echo=TRUE}
x <- c(TRUE, FALSE, TRUE, FALSE)
y <- c(TRUE, TRUE, FALSE, FALSE)
x
y
```

* Applying the logical operator `&` to `x` and `y` will apply an **element-by-element** 
logical **and** to the elements of `x` and `y`.

* That is, running `x & y` will return the following result
```{r, echo=TRUE, eval=FALSE}
x & y  ## & returns a vector comparing element-by-element
```

* Similarly, applying the logical operator `|` to `x` and `y` will apply an **element-by-element** 
logical **or** to the elements of `x` and `y`.
```{r, echo=TRUE, eval=FALSE}
x | y  ## | returns a vector comparing element-by-element
```

* Using the logical operator `!` with a vector will just return a vector where the `TRUE` values have been switched to `FALSE`
and the `FALSE` values have been switched to `TRUE`:
```{r}
!x
```


### && vs. & and || vs. |

* I would suggest using `&` for the logical **AND** operator and `|` for the logical **OR** operator.

* You may sometimes see `&&` and `||` being used in **R** code.

* `&&` and `||` can only be used for comparing vectors of length 1.
 
    - For vectors of length 1, they do the exact same thing as `&` and `|`

* For example,
```{r, echo=TRUE}
TRUE & FALSE  ## Same as TRUE && FALSE
TRUE & FALSE  

TRUE | FALSE  ## Same as TRUE || FALSE
TRUE || FALSE  
```


## If and If-else statements 


### if statements

* In **R**, the form of an [**if statement**]{style="color:#D96716"} is the following:
```{r, echo=TRUE, eval=FALSE}
if( condition ) {
    code_chunk1
}
```

* `condition` is usually a **logical expression**, but could just be a logical vector of length 1 (i.e., TRUE or FALSE). 

* If `condition` evaluates to **TRUE**, `code_chunk1` will be executed.

* You actually do not have to **indent** the code in `code_chunk1`, but I would **recommend** that you do indent.


* The code inside {...} will be executed 
only if the **condition** of the **if statement** is TRUE.

```{r, echo = TRUE}

if (TRUE) { # if condition is TRUE 
  "hello"   # this statement will run
}

if (FALSE) { # if condition is FALSE 
  "world"   # this statement will NOT run
}
```


### if statement examples

* Example 1: Running the following code will output the message in the `if` statement because the logical expression `x < y` evalutes to TRUE
```{r, echo = TRUE}

x <- 1
y <- 2
if (x < y ) {
  "x is smaller than y"
}
```

* Example 2: Running the following code will not print out anything:
```{r}
x <- 1
y <- 2
if (x > y ) {
  "x is greater than y"
}
```


* Example 3:
```{r, echo = TRUE}

x <- 3
y <- 2
if (x < y ) {
  "x is smaller than y"
}

if (x > y ) {
  "x is greater than y"
}
```


* Example 4:
```{r, echo=TRUE}
if( 2 < 3 ) {
    "Hello"
}
```

* Example 5:
```{r, echo=TRUE}
if( "dog" == "cat" ) {
    "Hello"
}
```

* Example 6:
```{r, echo=TRUE}
d = 2
if( d < 3 && d == 2.5) {
    "Hello"
}
```

* Example 7:
```{r, echo=TRUE}
if( 2 < 3 || 2 == 2.5) {
    "Hello"
}
```


### Single-line if statements

* If the code to be executed in the if statement is short, 
you can write it immediately after `if(condition)` on the **same line**.

* Or, you can write the single-line statement on the line immediately below `if(condition)`

```{r, echo=TRUE}
x = 5
if(x > 4 & TRUE) x = 2*x # multiply x by 2
x
```

* This **single-line** if statement is the same as using:
```{r, echo=TRUE}
x = 5
if(x > 4 & TRUE) { 
    x = 2*x # multiply x by 2
}
x
```


## if-else statements

* In many cases, you want to perform **an action** if a condition is true but perform **another action** if that condition is false.

* This can be done with an **if-else** statement.
* In **R**, the form of an **if-else** statement is the following:
```{r, echo=TRUE, eval=FALSE}
if( condition ) {
    code_chunk1
} else {
    code_chunk2
}
```

* As with if statements, `condition` is usually a logical expression, but could just be a logical vector (with a single element). 

* If `condition` evaluates to **TRUE**, `code_chunk1` will be executed.
    
* Otherwise, if condition evaluates to **FALSE**, `code_chunk2` will be executed.

---

* As an example, let's write an if-else statement that computes the **absolute value** of a number.

```{r, echo=TRUE}
x <- -3.2
if( x > 0 ) {
   abs_x <- x   # assign the variable abs_x the 
                # value stored in x
} else {
   abs_x <- -x   # assign the variable abs_x the 
                 # negative of the value stored in x
}
abs_x   # print the value stored in abs_x
```

* Another if-else example:

```{r, echo=TRUE}
x <- 5
if( x%%2 == 0 ) { # arithmetic operation x%%2 evaluated first
   "Hello"
} else {
   "world"
}
```

* Another if-else example:

```{r, echo=TRUE}
x <- 5
if( x%%2 == 0 | x > 4) {
   "Hello"
} else {
   "world"
}
```


### if-else-if chains

* In many cases, a desired computation will depend on **more**
than 2 conditions.

* For these cases, you can use an **if - else if - else** chain of conditional statements.

* The general syntax for an **if - else if - else** chain in **R** is:
```{r, echo=TRUE, eval=FALSE}
if ( condition1 ) {  ## If condition1 is TRUE,
    code_chunk1     ##  code_chunk1 will be executed.
} else if ( condition2 ) { ## If condition1 is FALSE
                           ## and condiiton2 is TRUE,
    code_chunk2          ## code_chunk2 will be executed
} else if ( condition3 ) { ## If both conditions 1 and 2 are FALSE
                           ## and condition3 is TRUE,
   code_chunk3             ## code_chunk3 will be executed
}
.
.
} else {      ## If all previous conditions are FALSE,
   else_chunk  ##  else_chunk will be executed
}
```


* An if-else if-else example:

```{r, echo = TRUE}

x = 2
if ( x < 0 ) {        ## If the condition is TRUE,
  "x is negative"     ## this statement will run.
} else if ( x == 0 ) {## If previous conditions are FALSE but this
  "x is zero"         ## is TRUE, this statement will run.
} else {              ## If previous conditions are FALSE,
  "x is positive"     ## this statement will run.
}

```


* Another if-else if-else example:

```{r, echo = TRUE}
message <- "second"
if ( message == "first" ) {   
    "hello"
} else if ( message == "second" ) {
    "world"
} else {
    "nothing"
}
```

---

* Be careful about the **location** of **else** in if-else if-else statements

* In **R**, you do not want to start a line with **else if** or **else**.


### Nested if-else statements 

* You can certainly have if-else statements **within**
a conditional statement.

```{r, echo = TRUE}
x = 3
if ( x %% 2 == 0 ) {  ## first condition
  if ( x < 0) {       ## second condition 
    "x is even and negative"
  } else {
    "x is even and non-negative"
  } 
}  else if ( x < 0 ) {## this if statement is not nested
  "x is odd and negative"         
} else {              ## not nested either
  "x is odd and non-negative"     
}

```


## Exercises

1. What will be printed to the screen if you run the **R** code below?
```{r, echo=TRUE, eval=FALSE}
x <- 2
if(3 < 2 || TRUE) {
    x <- 2*x
} else {
    x <- 0
}
print(x)
```




