# (PART) R Programming Fundamentals {-}

# Logical Expressions and If-Else Statements in R {#ifelse}



## Logical Expression in R

* A **Logical** expression is an expression that evaluates to either **TRUE** or **FALSE**.

* The following are examples of logical expressions in **R**:
    + `4 > 2`
    + `3 <= 5`
    + `15.0 + 1.3*1.3 > 17.0`
    + ` "cat" == "dog" `
    
* Each of the above expressions will evaluate to either
**TRUE** or **FALSE** if you run them in R.

```{r, echo=TRUE}
4 > 2
```

```{r, echo=TRUE}
3 <= 5
```

```{r, echo=TRUE}
15.0 + 1.3*1.3 > 17.0
```

```{r, echo=TRUE}
x <- "cat" == "dog" # assign to the variable x the value 
                    # returned by this logical expression
x
```

---

* Most logical expressions are constructed by using some **combination** of:
    + **Comparison** operators (<, <=, ==, !=)
    + **Logical** operators (**and**, **or**, **not**) (in **R**: &&, ||, !)

### Comparison Operators

```{r, echo=FALSE}
add.operator <- c("<", ">", "<=", ">=", "==", "!=")
add.meaning <- c("Less than", "Greater than", "Less than or equal to","Greater than or equal to","Equal to","Not equal to")
add.example <- c("5 < 3", "5 > 3", "3 <= 6", "4 >= 3","2 == 2","'str' != 'stR'")
add.result <- c("FALSE", "TRUE", "TRUE","TRUE","TRUE","TRUE")
AddOperators <- data.frame(Operator=add.operator, Meaning=add.meaning, Example =add.example, Result = add.result)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 4), padding=4)
```

### Logical Operators

* The first main logical operator we will discuss is the **logical AND**

* In **R**, the logical operator **&&** is used to represent the logical **AND**

* The **logical AND** is used to test whether or not two statements are **both** true.

* For two logical expressions **A** and **B**, the logical expression **A && B** is true
only if both **A** and **B** evaluate to true.

```{r, echo=TRUE}
4 > 2 && 5/2 == 1   ## only the first statement is TRUE
4 > 2 && "car" == "truck"   ## only the first statement is TRUE
4 > 2 && 3 < 5   ## both statements are TRUE
```

