# Loops in R {#loops}

* **Loops** are typically used when we want to repeat some type 
of calculation many times.

* The two types of loops in **R** are the **for** loop 
and the **while** loop.

* **For** loops: 
    + Typically used when we know exactly how many times we need to repeat a calculation.
    
* **While** loops:
    + Typically used when we only want to repeat some calculation until some condition is satisfied.


## For Loops

* The general form of a **for** loop in **R** is
```{r, echo=TRUE, eval=FALSE}
for (x in vec_name) {
    perform a calculation (often involving x)
}
```
    
* The **for** loop will execute the code underneath the **for** statement $T$ times where $T$ is the **length** of the vector `vec_name`.

* **Each** time the code is executed, `x` will be set to one element in `vec_name`.


### Example 1: printing the first 5 integers

* As an example, let's write a **for** loop that prints the first 5 positive integers:
```{r, echo=TRUE}
for (k in 1:5) {
    print(k)
}
```

* **Each time** you go inside the loop, the variable `k` takes the next value in the vector `1:5` and is printed to the screen.


### Example 2: computing 10 factorial

* As another example, let's use a **for** loop to compute the product
\begin{equation}
10! = 1 \times 2 \times 3 \times ... \times 9 \times 10 \nonumber
\end{equation}

```{r, echo = TRUE}
prod_ten <- 1   # start from 1

for(x in 2:10)  {  # iterate x from 1 thru 10
   prod_ten <- prod_ten*x   # multiply each value
}
print(prod_ten)     # print the product of first 10 integers
```


### Example 3: a non-numeric looping variable

* The looping variable **does not** need to be **numeric**.

* Here is an example where the looping variable has **character** type.
```{r, echo = TRUE}
fruits <- c("apple","orange","grape")   # character vector
for(fruit in fruits)  {  # iterate each element
   print(fruit)   
}
```


### Example 4: first Fibonacci numbers 

* The **Fibonacci sequence** is a sequence of numbers $x_{1}, x_{2}, x_{3}, ...$ defined by the relationship:
\begin{equation}
x_{k} = x_{k-1} + x_{k-2} \nonumber
\end{equation}

* The first two numbers in the Fibonacci sequence are $x_{1} = 0$ and $x_{2} = 1$.

* We want to write **R** code that stores the first $n$ Fibonacci numbers in a **vector**.

* To do this, it is more efficient to **first create** a vector of zeros and then fill in the Fibonacci numbers into this vector.
    + This is done by first assigning `x` the vector `rep(0, n)`

```{r, echo=TRUE}
## R code for computing first 12 Fibonacci numbers
x <- rep(0, 12)  # First initialize a vector of size 12
x[1] <- 0
x[2] <- 1
for(k in 3:12) {
    ## start at index 3 since first 2 elements are filled in
    x[k] <- x[k-1] + x[k-2]
}
x
```

---

* You could create a vector of the first $n$ Fibonacci numbers
by **"appending"** an extra number to the vector **x** each time
we go in the loop.

* This is usually much slower than first initializing a vector as was done 
on the previous slide.

```{r, echo=TRUE}
## R code for computing first 12 Fibonacci numbers
x <- c(0, 1)
for(k in 3:12) {
    x <- c(x, x[k-1] + x[k-2])
}
x
```


### Nested for loops

* You can have a for loop **within** another **for** loop. 
    + These are usually referred to as **nested** for loops.

```{r, echo = TRUE}
for (i in 1:3) {   # i iterates over 1,2,3
  for (j in 1:i) { # j iterates over 1 to i
      print(c(i,j))     
  }
}
```


### break and next: Exiting for loops and skipping iterations

* You can use the **break statement** to terminate a loop.
     + The **for** loop will be stopped whenever the statement **break** is encountered.
     
* As an example of this, let's write a loop that should run for 5 iterations, but is terminated after during it's 3rd iteration.

```{r, echo=TRUE}
for(k in 1:5) {
    print(k)
    if(k >= 3) {
       break   ## loop will stop when break is run
    }
}
```

* Note that if we move the `print` statement after `break` in the above example,
then only `1` and `2` will be printed to the screen.
```{r, echo=TRUE}
for(k in 1:5) {
    if(k >= 3) {
       break   ## loop will stop when break is run
    }
    print(k)
}
```
---

* **break** statements can be used within **nested loops**.

* If using a **break statement** in the **inner loop** of a nested loop, only the inner loop will be terminated when **break** is run.
```{r, echo = TRUE}

for (i in 1:3) {   # i iterates over 1,2,3
  for (j in 1:3) { # j iterates over 1,2,3
     print(c(i,j))
     if (j == i) {  ## if j and i are the same
        break        ## finish the inner loop 
     }
  }
}
```

---

* The **next statement** is used to skip the remainder of a current iteration.

* When the **next statement** is encountered, you will go directly to the next iteration and not execute whatever was remaining inside the body of the **for loop**.

* Let's try an example where we have 5 iterations setup, but we use a **next statement** within the 3rd iteration
```{r, echo=TRUE}
for(k in 1:5) {
   if(k == 3) {
       next
   }
   print(k)  # this will not print when k==3 
}
```


---

* `next` can also be used within a **nested loop**.

* If used in an **inner loop**, the **next statement** will skip the remainder of the current **inner loop** iteration and go to the next **inner loop** iteration.

```{r, echo = TRUE}
for (i in 1:3) {   # i iterates over 1,2,3
  for (j in 1:3) { # j iterates over 1,2,3
    if (j == i) {   ## if j equals i 
       next         ## go to next iteration
    }               ## of inner loop
    print(c(i,j))
  }
}
```


## While Loops

* **For** loops execute a piece of code a set number of times.

* **While** loops keep executing the piece of code as long as 
a certain condition is `TRUE`.

* The general form of a **while** loop in **R** is
```{r, echo=TRUE, eval=FALSE}
while(condition) {
    code chunk
    then, usually update something related to condition
}
```

* `condition` is a **logical** expression.

* The `code_chunk` inside the loop is repeated until the `condition` becomes `FALSE`
in which case the loop **stops**. 

---

* Basic example of while loop:
```{r, echo = TRUE}

prod <- 1   # start from 1
while(prod < 100)  {  # repeat the block until condition is true
  print(prod)          # print the value of prod
  prod <- prod * 2     # and double the value
}
    
```


### Example 1: a "first which" function

* We can use a **while loop** to find the **first occurrence** of a number in a vector.

* Specifically, for an input vector **x** and input number **num**, we want 
to find the first index `k` such that `x[k] == num`

* This function will also return `NA` if no match is found.
```{r, echo=TRUE}
first_while <- function(x, num) {
  index.count <- 0
  match.not.found <- TRUE
  while(match.not.found && index.count < length(x)) {
    # want to keep loop running while match is not found
    # and the index is less than the length of x
    index.count <- index.count + 1 
    match.not.found <- !(x[index.count] == num)
  }
  if(match.not.found) {
    index.count <- NA
  }
  return(index.count)
}
```

* Let's try running `first_while` on a few examples:
```{r, echo=TRUE}
first_while(1:5, 3)
first_while(1:5, 6)
first_while(1:5, 5)
first_while(rep(1:3, each=4), 2)
```

## Exercises

1. What will the value of the variable `count` be after running the following code: 
```{r, echo=TRUE, eval=FALSE}
x <- rep(c(-1,1), each=3)
y <- x > 0
x <- c(3, 3, 3, 5, 5)
count <- 0
for(k in 1:4) {
   count <- count + 1
   if( x[k] != x[k+1]) {
        break
   } 
}
```
