# Matrices in R {#matrix}

* A **matrix** is a table of data organized into **rows** and **columns**. 
\begin{equation}
\mathbf{A} = \begin{bmatrix} a_{11} & a_{12} & \ldots & a_{1p} \\
a_{21} & a_{22} & \ldots & a_{2p} \\
\vdots & \vdots & \ddots & \vdots \\
a_{n1} & a_{12} & \ldots & a_{np}
\end{bmatrix} \nonumber
\end{equation}

* The matrix $\mathbf{A}$ has $n$ rows and $p$ columns.
    
* The matrix element $a_{ij}$ represents the number in the $i^{th}$ row and $j^{th}$ column.

* **R** allows the elements of a matrix to be **any type**.

* As with vectors, the **elements** of a matrix in **R** must all have the **same type**.


## Creating matrices in R

* The most common way to create a matrix is 
with the **matrix** function.

* The form of the **matrix** function is
```{r, echo=TRUE, eval=FALSE}
matrix(x, nrow, ncol)
```

* **x** is a **vector** that you want to convert into a matrix.
    + `nrow` is the number of rows you want the matrix to have.
    
    + `ncol` is the number of columns you want the matrix to have.

    + `x` could be a single number (a vector of length one) in which 
      case the returned matrix will have all the same entries. 


* If we wanted to create a matrix named `A` with **2 rows** and **3 columns** that contains
**only zeros**, we could use the following code
```{r, echo = TRUE}
A <- matrix(0, 2, 3) 
A
```

* To **convert** a vector into a matrix 

* **Convert** the vector (1, 2, 3, 4, 5, 6) into a matrix with 2 rows and 3 columns
```{r, echo=TRUE}
x <- 1:6     # x is a vector of 1,2,3,4,5,6 
B <- matrix(x, 2, 3) # create a matrix from this vector
B
```


* Notice that **matrix** function fills in the matrix entries by columns (i.e, it fills in the numbers from `x` in the first column, then moves to the second, etc.)
   + This is referred to as filling in the entries "**by column**"
```{r, echo = TRUE}
x <- 1:6     # x is a vector of 1,2,3,4,5,6 
B <- matrix(x, 2, 3) # create a matrix from this vector
B
```

* You can instead fill in the matrix entries **by rows** by using the **keyword** argument `byrow = TRUE`
```{r, echo=TRUE}
D <- matrix(x, 2, 3, byrow=TRUE)
D
```

---


* It is often useful to create a matrix which has **all the same** columns (or rows):
```{r, echo = TRUE}
matrix(1:2, 2, 3) # duplicates the column vector c(1,2)

matrix(1:3, 2, 3) # this does not duplicate the row vector c(1,2,3)

matrix(1:3, 2, 3, byrow=TRUE) # but this one does!
```

## Accessing different features of a matrix

* To access specific **elements** of a matrix:

```{r, echo = TRUE}

A <- matrix(1:6, 2, 3, byrow=TRUE) # creating a 2x3 matrix
A                         # print the matrix

A[2,1]
A[2,1:2]
A[,2]
```



* Using **nrow()** or **ncol()** let you see the number of 
**rows** or **columns** of a matrix.

* **dim** will return both the number of rows and columns.

```{r, echo = TRUE}

A <- matrix(1:6, 2, 3, byrow=TRUE) # creating a 2x3 matrix

nrow(A)
ncol(A)
dim(A)
```

---


* You can also access the elements of a matrix as you would with a **vector**. 

* For a matrix, the indexing increases column-by-column.
    + For example, if `A` has 3 rows: 
    + `A[1]=A[1,1]`, `A[2]=A[2,1]`, `A[3]=A[3,1]`, `A[4]=A[1,2]`, ...

* I don't really use this way of accessing matrix elements much though, 
in some cases, it is more convenient (e.g., `A[ which(A > 3) ]`)

```{r, echo = TRUE}
A 
A[4:5]          ## accessing like a one-dimensional vector
```

* Accessing elements of a matrix like a vector using **logical subsetting**

```{r,echo=TRUE}
which( A > 3)      ## getting indices with which
A[ which(A>3) ]   ## getting the elements greater than 3
```



* The **diagonals** of a matrix are the elements of the matrix where
the **row index** equals the **column index**.

* In a **diagonal matrix**, all **non-diagonal** elements must be zero.

* To create a **diagonal** matrix, use the `diag` function.
    + You must provide the vector of **diagonal elements** as the input to `diag`
```{r, echo = TRUE}
diag(1:4)   # 4x4 diagonal matrix with (1,2,3,4) as diagonals
```


* If you use `diag(A)` where the **input** `A` is a **matrix**, this 
will return the diagonal elements of `A`.

```{r, echo = TRUE}
A <- matrix (1:9, 3, 3)
A
diag(A)   # returns diagonals of this matrix
```


* Working with **off-diagonal** elements 

```{r, echo = TRUE}
x <- matrix (1:9,3,3)
x[upper.tri(x)] <- 0
x
x[lower.tri(x,diag=TRUE)] <- 10:15
x
```

### Naming rows or columns

* You can give a name to each row with the **rownames** function.
    + The collection of row names of a matrix should be a vector.

```{r, echo = TRUE}
x <- matrix(1:6, nrow=2, ncol=3)
rownames(x) <-c("a","b")
x
```

* You can actually access rows of a matrix **by name**:

```{r, echo = TRUE}
x[1,] 
rownames(x) <-c("a","b")
x["a",]
```

* Similarly, you can access **columns** of a matrix **by name**:

```{r, echo = TRUE}
colnames(x) <-c("c1","c2","c3") 
x
x[2, c(2,3)]
x["b", c("c2","c3")]
```

### Combining rows or columns

* You can **"join"** matrices in "vertical" or "horizontal" ways
by using the functions 
   + `cbind` (**column** bind)
   + `rbind` (**row** bind)

```{r, echo = TRUE}
X <- matrix(1:6,2,3)
Y <- X^2
cbind(X,Y)
rbind(X,Y)
```



## Applying functions to matrices

* Functions like **sum** or **mean** will
take the sum (or average) of **all** 
the elements of a matrix:

```{r, echo = TRUE}
A <- matrix (1:6, 2, 3)
sum(A)  # sum of all elements
mean(A) # average of all elements
sd(A)   # standard deviation of all elements
```


* Useful operations on matrices

```{r, echo = TRUE}
A
t(A)             ## transpose of A (reverse role of rows/columns)
as.vector( A )   ## convert matrix to vector in column-wise order
as.vector( t(A) ) ## convert matrix to vector in row-wise order

```

### Mathematical operations with matrices

* Using `+,-,*,/` with two matrices performs **element-wise** addition,
subtraction, multiplication, and division.

```{r, echo = TRUE}
y <- x <- matrix(1:6,2,3,byrow=TRUE)   
x + y  # matrix addition (element-by-element)
x - y  # matrix subtraction (element-by-element)
x*y  # element-wise multiplication
```


* Matrix multiplication is done with `%*%` 

```{r, echo = TRUE}
x %*% t(y)  # (2x3) x (3x2) matrix
t(x) %*% y  # (3x2) x (2x3) matrix
```



## Matrix Functionals

### Functions applied to **each** row/column

* We can take the **sum** or **mean** of each row/column of a matrix.

```{r, echo = TRUE}
A
rowSums(A)
colSums(A)
rowMeans(A)
```


* Writing our own **rowSums()** function

```{r, echo = TRUE}
myRowSums <-function(x){
  ret <- rep(NA,nrow(x)) ## make an empty vector
  for(i in 1:nrow(x)) {
    ret[i] <-sum(x[i,])  ## take the sum of i-th row
  }
  return (ret)
}
rowSums(A)
myRowSums(A)
```


* The implementation by the function `myRowSums` works fine.

* However, it is a bit cumbersome to write such a function
whenever we want to compute **column-wise/row-wise** summary statistics.

* It would be nice if there were a **quick & general** way to do the following:
  + For each row or column of a matrix ...
  + Evaluate a function using the row/column vector as input
  + Return the collection of results as a numeric vector.
  

### **apply()**: dimension-wise aggregation


```{r, echo = TRUE}
rowSums(A)    ## original row-wise sum
apply(A, 1, sum) ## this also performs row-wise sum
colSums(A)    ## original column-wise sum
apply(A, 2, sum) ## this also performs column-wise sum
```



* **First parameter** of apply: the **input matrix**.

* **Second parameter** of apply: the **dimension** over which to aggregate 
    + 1:rows,  2:columns

* **Third parameter** of apply: **function** to evaluate for each row/column

```{r, echo = TRUE}
A
apply(A, 1, sum)  # row sums
apply(A, 2, sum)  # column sums
```


* Examples of using apply:

```{r, echo = TRUE}
colMeans(A)  ## original columnw-wise mean
apply(A, 2, mean) ## same as above
# colMax(A)   ## This will cause ERROR. colmax() does not exist!
apply(A, 2, max) ## But this works!
```



### Using apply with your own function

* Suppose that we want to do the following:

* For **each** row/column:
  + Calculate the squared sum of each row/column.
  + Return the vector of squared sums.

* How can we do this using **apply()**?

---


* A **big** advantage of the **apply** function is the
**flexibility** it gives you.

* You can use **your own** functions to compute **row-wise/column-wise summary measures** 
for certain measures that may not be implemented in base **R**. 

```{r, echo = TRUE}
## define a function named sqsum
sqsum <- function(x) {
  return( sum(x*x) )  ## very simple implementation
}
apply(A, 1, sqsum)    ## run sqsum() funciton
```

---


* You can actually write the function definition **inside**
of the apply function.

* This can be convenient when using straightforward functions that can be written on a single line:
```{r, echo = TRUE}
apply(A, 1, function(x) {return (sum(x*x))} )    ## much simpler!
```

* Or, we can write the above apply statement in an even simpler form:

```{r, echo = TRUE}
## You can omit return() and {} especially 
## for simple function definitions
apply(A, 1, function(x) sum(x*x) )    ## Even simpler!
```

---

* **apply()** can also return a **matrix**

```{r, echo = TRUE}
## if apply() returns a vector, each row stores results
apply(A, 1, function(x) c(min(x),max(x)) )
apply(A, 2, function(x) c(min(x),max(x)) )
```


## Exercises

1. Write a **function** that has two input arguments **A** and **B**
    + **A** and **B** are both matrices with the same dimension (same number of rows and columns).
    + The function returns a matrix **D**.
    + The `[i,j]` element of **D** should equal `A[i,j]` if `A[i,j] < B[i,j]` and equal `B[i,j]` if `A[i,j] >= B[i,j]`.
    + You can use either a **nested** for loop or **logical indexing**.
    
2. Suppose we define the matrix `X` as
```{r, echo=TRUE, eval=FALSE}
X <- rbind( rep(c(1,4), each=3), rep(c(8,6), each=3))
```
What will then be the value of `apply(X, 2, sum)[2]`?


    
    
    
