# Lists {#list}

* R always store data as a **"collection"**.

```{r, echo=FALSE}
add.dimension <- c("1-Dimension", "2-Dimension", ">2-Dimensions")
add.homogeneous <- c("Atomic Vector", "Matrix", "Multi-dimensional array")
add.heterogeneous <- c("List", "Data Frame", "")
AddOperators <- data.frame(Dimension=add.dimension, Homogeneous=add.homogeneous, Heterogeneous=add.heterogeneous)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 3), padding=3)
```
* There is no "0-dimensional data" in R.

* Even a single-valued object is considered to be a **"vector"** with **length 1**.

Source: http://adv-r.had.co.nz/Data-structures.html 


## Creating Lists in R

* **Lists** can store a collection of items that have **different types**.

* To **create a list**, you can use the `list` function. 

```{r, echo=TRUE}
## Use a keyword argument to define any attribute of any data type
emp_info <- list(name="Nicholas Henderson", UMID=12345678, 
                 faculty=TRUE)
emp_info # list has character, numeric, and logical types
emp_info$name # access components with $component_name
```


* Create a list with **named components**: **a**, **b**, and **d**.

```{r, echo=TRUE}
list(a=c(1,2,3), b=c("apple","banana"),
     d=TRUE) ## list elements can be vectors
```


## Accessing list elements

* You can create **named lists** or **unnamed lists**.
    + In an **unnamed list**, the list components **do not** have names.

* You can access the $k^{th}$ element of an **unnamed list** using 
the double bracket syntax `[[k]]`.

---

* You can access the elements of a **named list** using either the 
indexing `[[k]]` or the component names.

```{r, echo=TRUE}
# Create an unnamed list and look at the first element:
y <- list(c(1,2,3), c("apple","banana"), TRUE)  
y[[1]]  
```

---

* Accessing the elements of a named list using 
either the **numeric index** or the **component names**:
```{r, echo=TRUE}
x <- list(a=c(1,2,3), b=c("apple", "banana"), d=TRUE) 
x$a
x[[1]]
x$b
x[[2]]
```


## Working with lists

* Get the names of the list components by using `names(list_name)`
```{r, echo=TRUE}
names(x)
```

* Find the length of a list by using `length`.
    + Note that the length of a list is the **number of components** of the list,
    and the length does not depend on the length of any of the vector components.
    
```{r, echo=TRUE}
length(x)
```



### Expanding a list: adding more components

* Add components to a list using either the index or by name
```{r, echo=TRUE}
x
x[[4]] <- 4
x$e <- 5
```

```{r, echo=TRUE}
x
```


##  When are lists useful?

* Many **R** functions return **multiple** items as output. 
    + It is common to return these multiple items in a single, **named list**. 
    
```{r, echo=TRUE}  
matrixSum <- function(X) { # calculate possible sums of a matrix
  s1 <- rowSums(X)
  s2 <- colSums(X)
  s3 <- sum(X)
  return(list(row=s1, col=s2, all=s3)) ## return multiple values 
                                       ## using a list
}

A <- matrix(1:6,2,3)
matrixSum(A)
```



* Lists are **useful** in situations where you have a collection of items 
of **different types** and **dimensions** that you want
to store in a single **"variable"**.

* For example, you may have data on several individuals
where, **for each individual**, you have data stored 
in a vector, a matrix, and a data frame.



## lapply(): Applying functions to lists

* **lapply** ("list apply") works like an **apply()** function
for **lists**.

* **lapply** has two arguments:
    + An input **list**.
    + The **name** of the **function** to apply to each component of the list.

* **lapply** returns a list with the **same number** of components as the input list:
```{r, echo=TRUE}
r <- matrixSum(A) # r is a list with 3 components.
lapply(r, length) # compute the length of each component of r
```


```{r, echo=TRUE}
lapply(r, sum) # Compute sum of each component of r
lapply(r, function(x) sum(x*x)) # Compute sum of squares for each component
```


### Use sapply() for simpler output

* When applying a function to **each** component of a list,
it is often preferable to have the **result** returned as a **vector**
or **matrix** instead of a list.

* This is what the **sapply()** ("simplified lapply") function does.

```{r, echo=TRUE}
sapply(r, sum) # sapply() returns output as a named vector
lapply(r,sum)
```


* If the **function** used in **sapply** returns a vector 
with length **longer** than 1, **sapply** will return a matrix.

```{r, echo=TRUE}
r
                                ## column names of matrix are same
sapply(r, function(x) c(min(x),max(x)) ) ## as list component names
```



## Exercises

1. Suppose we define the vector `x` as `x <- c(1,2,3)`. What is the value of 
```{r, echo=TRUE, eval=FALSE}
length( list(x) ) == length(x)
```

2. Write a function that takes a **list** called **x** as input, and the elements of `x` can be assumed to be numeric vectors. The function should **return** a matrix where
    + the **number of columns** of the matrix equal the number of elements of the input list.
    + the **number of rows**of the matrix equals 4
    + If the $k^{th}$ element of the list has length greater than 4, 
the $k^{th}$ column of the output matrix should be `(mean(x[[k]]), median(x[[k]]), min(x[[k]]), max(x[[k]]))`
    + If the $k^{th}$ element of the list has length 4 or less, the 
$k^{th}$ column of the output matrix should be `(0, 0, 0, 0)`.




