# Data Frames {#dataframe}


```{r, echo=FALSE}
add.dimension <- c("1-Dimension", "2-Dimension", ">2-Dimensions")
add.homogeneous <- c("Atomic Vector", "Matrix", "Multi-dimensional array")
add.heterogeneous <- c("List", "Data Frame", "")
AddOperators <- data.frame(Dimension=add.dimension, Homogeneous=add.homogeneous, Heterogeneous=add.heterogeneous)
knitr::kable(AddOperators, format = 'markdown', align=rep('c', 3), padding=3)
```

* **Data Frames** are the most common tool for storing "rectangular" datasets (data stored as a table in rows and columns) in **R**.
    + Each row holds an **observation**.
    + Each column holds a **variable**.

* Roughly speaking, a data frame can be thought of as a matrix where 
**different columns** can hold **different types** of data.
    + The elements in the **same column** of a data frame should have the **same type**. 


* More technically, a data frame is actually a .red[list of vectors] that all 
have the same length. **Number of variables = number of elements in list.**

---

* Let's first look at one of **R's** built-in datasets to see an example of a data frame.

* **R** has many **built-in datasets** that are stored as data frames.

* Type in `library(help = "datasets")` to see the full list of built-in
datasets.

* One of the built-in datasets is the **PlantGrowth** dataset.

* You can **load** this dataset into your **R** session by typing in
```{r, echo=TRUE}
data(PlantGrowth)
```

* Check that `PlantGrowth` is a data frame by using
```{r, echo=TRUE}
class(PlantGrowth)
```

---

* You can display the first few rows of the **PlantGrowth** data by using `head()`:
```{r, echo=TRUE}
head( PlantGrowth )
```

* This dataset has **30 observations** and **2 variables**.
    + This can be seen by applying `dim()` to the data frame: 
```{r, echo=TRUE}
dim( PlantGrowth )
```


## Creating Data Frames

* The main way to **create** a data frame in R is to use the function `data.frame`.

* To create a data frame with 5 observations and two variables named **x** and **y**:
```{r, echo=TRUE}
# x and y have to be the same length
df <- data.frame(x=c(1,2,3,4,5), y=c(1,4,9,16,25))
df
```

* **Converting** a data frame to a matrix

```{r, echo=TRUE}
df.mat <- as.matrix(df) ## How a Data Frame looks like  
df
df.mat  ## How a matrix looks like
```


* You can use `as.data.frame` to convert a matrix into a data frame. 

```{r, echo=TRUE}
X <- matrix(1:6,3,2) 
colnames(X) <- c("a","b")
X.df <- as.data.frame(X)
X
X.df
```

* Remember that a data frame can take **heterogeneous** data types.
```{r, echo=TRUE}
df <- data.frame(x=c(1,2,3),y=c("a","a","bc"))
colnames(X) <- c("a","b")
X.df <- as.data.frame(X)
X
X.df
```


## Accessing and Modifying Data Frames

* One way to **access** the individual elements of a data frame is by specifying the row/column index. 
    + This is done in exactly the **same way** as with a **matrix**.

```{r, echo=TRUE}
df[,1]   ## access a column just like a matrix
df[1:2,] ## access rows like matrix
df[2, 1] ## access individual elements
```


* You can access the data from a specific variable **by name** 
using the syntax: `df_name$var_name`.

```{r, echo=TRUE}
df$y   ## access a column by name
       ## what happens if we omit a comma?
df[1]  ## (Remember that df is actually list)
```


* Differences in element access between matrices and data frames:
```{r, echo=TRUE}
df.mat <- as.matrix(df) 
df[2] #df[10] will not work
df.mat[2]
df.mat[10]
```


* **Modifying** data frames:

```{r, echo=TRUE}
df <- data.frame(x=c(1,2,3),y=c("a","a","bc")) 
df$z = c("def","123","123") ## adding a new column
df
df2 <- data.frame(w=c(TRUE,FALSE,FALSE))
cbind(df,df2)  # combine columns
```


* **Adding** new rows to a data frame:
```{r, echo=TRUE}
df3 <- data.frame(x=4,y="a",z="def") 
rbind(df,df3)  
```


## Subsetting data frames

```{r, echo=TRUE}
df
subset(df, x>1)
subset(df, x>1 & y =="a")  
```

## Exercises
