# Factors {#factors}


* **Factors** in **R** were created to represent **categorical variables** in statistics.

* **Categorical Variable:**
    + **Categorical variables** represent membership in some category rather than a numerical value **(e.g., blood type, country, language spoken, treatment group)**.
    
    + **Categorical variables** can only assume one of a finite collection values.
    
    + It does not usually make sense to make **numerical comparisons** between categorical variables.

## Definition of Factors

* In **R**, a factor can be thought of as a vector that
contains **specific extra information** about that vector.
    + More specifically, a factor is a vector that has certain **attributes** that
    distinguish from non-factor vectors.

* One key attribute stored by a factor vector is its **levels**. 

* The **levels** of a factor vector determine the **set of allowable values** in that vector.

* For example, let's look at the **group** variable in the **PlantGrowth** dataset 
```{r, echo=TRUE}
data(PlantGrowth)
PlantGrowth$group
```

* The **group** variable is a factor with the following 3 levels: 
**ctrl**, **trt1**, **trt2**.


* It is appropriate to make the **group** variable in **PlantGrowth** 
a **factor** since this variable represents the treatment condition in this study.

* Because the **levels** of group (**ctrl**, **trt1**, **trt2**) represent the collection of possible values,
you **cannot** assign the elements of **PlantGrowth$group** a value 
which is not either **ctrl**, **trt1**, or **trt2**.
```{r, echo=TRUE}
PlantGrowth$group[4] <- "trt" 
```

* Assigning the 5th element the value `"trt1"` works though:
```{r, echo=TRUE}
PlantGrowth$group[5] <- "trt1" 
```


---


* The other **main attribute** which distinguishes factors is its **class**.
```{r, echo=TRUE}
attr(PlantGrowth$group, "class")  # Check the class attribute
```

* If the **class** of a vector is a **factor**, numerical comparisons
between elements will return an error:
```{r, echo=TRUE}
PlantGrowth$group[5] > PlantGrowth$group[15]
```

```{r, echo=TRUE}
PlantGrowth$group[5] + PlantGrowth$group[15] # This returns an error
```


## Adding new levels to a factor

* In some cases, you want to **add data 
to a factor vector**, and the new data contains new
categories which are not in the original factor vector.
  
* To do this, you can assign **new levels** to your factor 
vector by using the assignment:
```{r, echo=TRUE, eval=FALSE}
`levels(vec_name) <- `
```

* For example, look what happens when we try to change the 
value of one element from a factor vector
```{r, echo=TRUE}
y <- factor(rep(1:3, 3)) # create factor with levels 1, 2, 3
y[3] <- 4  # Try to assign 3rd element the value 4
y  # Note that the third element is now NA
```


* We **can add** the value $4$ to this vector, but we need to change
the **factor levels** first:
```{r, echo=TRUE}
levels(y) <- c(1,2,3,4)
y[3] <- 4
y
```

---

* The `str()` function gives a description of any **R** object.

* For a data frame, you can see which variables are treated as **factors**
and which are treated as **numeric**

```{r, echo=TRUE}
str( PlantGrowth )
```

```{r, echo=FALSE}
data(PlantGrowth)
```



## The table() function

* The `table()` function can be used to count the **number of observations** at 
each **level** of a factor.

* `table()` can also be used to count the number of observations
ocurring at each .**level combination** for **2 or more factors**.

* For example, let's count the number of observations in each treatment 
group in the **PlantGrowth** data:
```{r, echo=TRUE}
table( PlantGrowth$group ) # 10 observations in each group
```

---

* To see how **table** works for multiple factors, let's look at
the **VA** dataset from the **MASS** package
```{r, echo=TRUE}
library(MASS)
data(VA)  # load the VA data frame
str(VA)  # look at the structure of this data frame
```


* Notice that the variables **treat** and **cell** are **factors** in the **VA** data frame.

* You can **cross-tabulate** these two factors by using plugging both variables into the **table** function.
```{r, echo=TRUE}
tab.VA <- table(VA$treat, VA$cell)
tab.VA
```

* **tab.VA** is an example of what's usually called a 2x4 **contingency table.**

* You can compute the table margins as if **tab.VA** were a matrix
```{r, echo=TRUE}
colSums(tab.VA)
```


* You are not limited to cross-tabulating two factor variables.

* For example, we can cross-tabulate the factors **treat**, **cell**, and **prior**:
```{r, echo=TRUE}
table(VA$treat, VA$cell, VA$prior)
```

* This outputs **two** 2x4 contingency tables: A contingency table of **treat** and **cell** for 
each level of **prior**.


## tapply() and aggregate

* `tapply()` is useful when you want to compute
certain summary statistics **for each** level of a factor variable.

* For example, suppose we wanted to compute 
the **mean** and **standard deviation** of 
**age** in the **VA** data within each treatment group.

* This can be done with the following code:
```{r, echo=TRUE}
tapply(VA$age, VA$treat, mean) # compute mean age in each trt. group
tapply(VA$age, VA$treat, sd)  # compute sd of age in each trt. group
```


* You can use the `aggregate()` function to 
use `tapply()` across multiple variables.

* This is useful, for example, if you want to compute
the mean of **each** of the numeric variables in a data frame
using a single command.

* Note that the variables **stime**, **status**, **age**, **Karn**,
**diag.time** in the **VA** data are numeric:
```{r, echo=TRUE}
str(VA)
```


* We can compute the mean of each of the numeric variables in 
**VA**, stratified by cell type, by using `aggregate()` in the following way:
```{r, echo=TRUE}
# Use aggregate on VA only using 1st, 2nd, 4th, 5th, 6th variables
aggregate(VA[,c(1,2,4,5,6)], list(VA$cell), mean)
```

* Compute the **median** of each of the variables stratified by cell type
```{r, echo=TRUE}
aggregate(VA[,c(1,2,4,5,6)], list(VA$cell), median)
```
