# (PART) Importing and Organizing Data {-}

# Reading Data Files into R {#fileinput}

* **Text files**

  + Files with human readable characters such as letters, numbers, special characters (e.g., .red[", ', ?]), typically composed of multiple lines.
  
  + Can represent different characters by text encodings 
  
  + There are many types of text files, e.g.,
      
      -R files (.R)
      
      -html files (.html) / cascading style sheet (.css)
      
      -text files (.txt)
    

## Downloading a file locally

* Need to download a file to your local computer

* To test out reading in datasets into **R** that you have stored locally on your computer, we will first download the **gapminder** dataset.

* When you download a file, make sure you know
which **folder** it is in and that you can find the file path for the downloaded file.
   

* To load the **gapminder** dataset into **R**, you can use the **read.csv** function.

* Before running `read.csv`, it is often **convenient** to first set the **working directory** in **R** first using `setwd`
   + This allows you to load in all the files from the folder containing your data files without typing in the full file path name every time
   
   + For example, if the **gapminder** data was stored in a folder with path **"~/IntrotoR/Data**, we could first type
```{r, eval=FALSE}
setwd("~/IntrotoR/Data")
```

* After running `setwd`, we can now load in the **gapminder** dataset (stored as `gapminder_full.csv` on my computer) using `read.csv`
```{r, echo = TRUE, eval=FALSE}
fname <- "gapminder_full.csv"
gapminder <- read.csv(fname)
```

```{r, echo = FALSE}
fname <- "~/Documents/Stat607Notes/Data/gapminder_full.csv"
gapminder <- read.csv(fname)
```

---


* To import a .csv file (or any other text file) in **Rstudio**, you can also click on the **Import Dataset** button on the top right panel.

* This lets you view how the data would appear before it is loaded into **R**.

---


* The **read.table()** function is the more general function 
for reading in rectangular data stored in text files.
   + You can read in other types of text files: e.g., **.tsv**, **.txt**

* **read.table()** can do the same thing as **read.csv()**. The main difference is that the default settings of **read.csv()** are setup to handle .csv files nicely.

* For example, we could read in the **gapminder** data with **read.table()** using:
```{r, echo = TRUE}
df <- read.table(fname, header=TRUE, sep=",") # Set header=TRUE
head(df)
```

---

* You can look at the **first few rows** of a loaded data frame using `head`:

```{r, echo = TRUE}
dim(gapminder) # 1704 observations, 6 variables
head(gapminder, 8)  ## look at first 8 rows
gapminder$country[1:5] # note that country is a factor
```

---

* Looking at the **structure** of the data frame using `str()`

```{r, echo = TRUE}
str( gapminder )
```

---

* Reading strings as **non-factors**: 

```{r, echo = TRUE}
gapminder_nofac <- read.csv(fname, stringsAsFactors = FALSE) 
head(gapminder_nofac)
gapminder_nofac$country[1:5] # No more "levels"
```

---

* You can **skip** lines when you read in a text file.
    + This is useful if you don't want to read
in the "header" line of the .csv file: 
```{r, echo = TRUE}
df <- read.table(fname, header=FALSE, sep=",",skip=1) 
head(df)
```

* Or, skip the header and the first two rows of the .csv file
```{r, echo = TRUE}
df <- read.table(fname, header=FALSE, sep=",",skip=3) 
head(df)
```


## Opening a remote file using read.table()

* You can read in a dataset directly from a URL:
```{r, echo = TRUE}
# Read in ignoring the fact that there is a header line
urlname <- paste("https://web.stanford.edu/~hastie/",
                 "ElemStatLearn/datasets/bone.data", sep="")
bone_dat <- read.table(urlname)
head(bone_dat, 4)
```

```{r, echo=TRUE}
# Set header=TRUE to read in top line as the variable names
bone_dat2 <- read.table(urlname, header=TRUE)
head(bone_dat2, 4)
```


