# Data Wrangling {#datawrangling}

* **Data Wrangling** just refers to 
the process of organizing your data into a useful format so 
that you can 
    + Perform the **statistical analysis** you are interested in.
    
    + Produce the **data visualizations** you are interested in.

* The amount of work involved in the data wrangling process will
vary a lot depending on how **"clean"** the original dataset is. 

* Examples of steps involved in data wrangling:
    + **subsetting data**,
    + **merging data**, 
    + **transforming data**, 
    + **deriving new variables**,
    + **handling missing data**

## The nycflights13 data

* To illustrate the techniques described in this chapter, we will use the **"NYC Flights Dataset"** available from the `nycflights13` **R** package.

* To load this package into **R**, first install the `nycflights13` and 
then load it into **R** with the following command:
```{r}
library(nycflights13)
```

* This package has several datasets available.

* We will explore the `flights` dataset in this chapter. 
    + This dataset can be loaded into **R** with the following command.
```{r}
data(flights)
```

* tibble vs. data frame here.
```{r}
flights <- as.data.frame( flights )
```

## Exploring the Dataset

* The first thing you usually want to note about a dataset after first loading it is **how many observations** and **variables** there are.

* While you can note the dimensions of your data under the **"Environment"** tab in Rstudio, you can see this by using `dim`
```{r}
dim(flights)
```

* This data frame has $336,776$ observations and $19$ variables.

* To see what is stored in the **first few rows** of `flights` use `head`:
```{r}
head(flights)
```

---

* You can use `str` to see which variables are stored as **numeric** vs. **character** vs. **factor**.
    + In many cases, you may want to change the character variables to factor.
    
    + You may notice that one of the variables in this dataset has the `POSIXct` type which is a special data type used to store times.

```{r}
str(flights)
```

---

* Each row of `flights` holds information related to a **single flight** that departed from the New York City area in 2013.

* The `carrier` variable in `flights` contains the abbreviation for the airline carrier. 

* We can see that there are 16 **different carriers** in this dataset by using `unique`:
```{r, echo=TRUE}
length( unique(flights$carrier) )  ## 16 unique carriers
```

* Each plane has a unique **tail number**. Using `unique`, we can see that 
there are $4044$ unique planes represented in the `flights` dataset
```{r}
length( unique(flights$tailnum) )
```

* Using `table`, we can see which **carriers** had the most flights out of the New York City area:
```{r}
table( flights$carrier )
```

## Summarizing each variable



## Using the dplyr package

